humanCdGF_predict_cfu

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
update reduced features for one mouse model

#One per cage
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))

om_rsq_rf <- c() # rsq of each rf model
om_rsq <- c() # cumulative rsq of all models
om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
om_pred <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    om_rsq_rf[i] <- tail(rf_om$rsq, 1)
    om_pred <- rbind(om_pred,
                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                          predicted=predict(rf_om, test)))
    om_rsq[i] <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_import_otus <- select(om_import_otus, -IncNodePurity)
}
test <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
colnames(test) <- seq(ncol(test))
### rank
test_rank <- apply(test, 2, function(x) rank(x))
test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
test_import_rank <- sort(test_raw_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
### mean
#test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
#test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
#test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
#test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
#### median
#test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
#test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
#test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
#test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)

#####
#test <- c() # test if any model has same train set
#  for(i in 2:21){
#  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
#}
#any(test) # if true, then trained on the same set of mice multiple times
#####

@@

med_otu_importance_om <- data.frame(otu = names(test_import_rank),
                                 importance = test_import_rank)

@@

## Feature reduction

#### Boruta  
Variance Explained = `r round(tail(rf_boruta$rsq, 1)*100, 2)`%  
OTUs  
`r rf_bor_import`

#### Mean % Increase MSE  
`r length(otus_used[[1]])` OTUs = Variance Explained = `r round(rf_rsq[[1]]*100, 2)`%  
`r length(otus_used[[2]])` OTUs = Variance Explained = `r round(rf_rsq[[2]]*100, 2)`%  
`r length(otus_used[[3]])` OTUs = Variance Explained = `r round(rf_rsq[[3]]*100, 2)`%  
`r length(otus_used[[4]])` OTUs = Variance Explained = `r round(rf_rsq[[4]]*100, 2)`%  
`r length(otus_used[[5]])` OTUs = Variance Explained = `r round(rf_rsq[[5]]*100, 2)`%  
    - OTUs:   `r otus_used[[5]]`  
`r length(otus_used[[6]])` OTUs = Variance Explained = `r round(rf_rsq[[6]]*100, 2)`%  
    - OTUs:   `r otus_used[[6]]`  
`r length(otus_used[[7]])` OTUs = Variance Explained = `r round(rf_rsq[[7]]*100, 2)`%  
    - OTUs:   `r otus_used[[7]]`

Either Enterococcus (Otu 5) or Clostridium_XlVa (OTU 19) alone allows for ~65% explained variance

#### one mouse model beginning with all otus and removing OTUs ranked below half max
Rsq 0.6505947 with all otus

Rsq 0.6893368 with otus:
Otu00001 Otu00002 Otu00003 Otu00004 Otu00005 Otu00006 Otu00009 Otu00011 Otu00012 Otu00014 Otu00015 Otu00016 Otu00017 Otu00018 Otu00019 Otu00021 Otu00022 
 19727.0   7883.0  10100.0  20087.0  19774.0  15910.0   6999.0  10213.5  13479.0  16050.0  19178.0  15146.0  11461.0   8879.0  21037.0   8616.0   8201.0 
Otu00029 Otu00030 Otu00032 Otu00045 Otu00048 Otu00049 Otu00053 Otu00057 Otu00060 Otu00063 Otu00064 Otu00074 Otu00077 Otu00081 Otu00082 Otu00083 Otu00084 
  9849.0  13750.0   8594.0   7568.0  14825.0  12015.0  12291.0  13443.0  10921.0   8801.0   7474.0   9516.0   8280.0   9914.0  11609.0  12119.0  12961.0 
Otu00095 Otu00101 Otu00118 Otu00128 Otu00131 Otu00159 Otu00163 Otu00164 Otu00173 Otu00175 Otu00176 Otu00178 Otu00199 Otu00200 Otu00250 Otu00297 Otu00461 
 15665.0  15639.0  10566.0  10740.0   9099.0  14344.0   9217.0  18500.0  11910.0   8002.5  19911.0   8550.5  20788.0  21936.0  22797.0  20767.0   7897.5

Rsq 0.7430718 with otus:
Otu00001 Otu00004 Otu00005 Otu00006 Otu00012 Otu00014 Otu00015 Otu00016 Otu00019 Otu00030 Otu00048 Otu00057 Otu00095 Otu00101 Otu00159 Otu00164 Otu00176 
    6855     6074     6119     5436     4061     4580     6285     4193     7725     3766     3350     3980     4305     4138     3282     5724     6645 
Otu00199 Otu00200 Otu00250 Otu00297 
    6845     7898     7511     6728

Rsq 0.7447479 with otus:
Otu00001 Otu00004 Otu00005 Otu00015 Otu00019 Otu00164 Otu00176 Otu00199 Otu00200 Otu00250 Otu00297 
    3081     2423     3019     3236     3517     2546     2521     2830     3151     3739     2937

Rsq 0.7572649 with otus:
Otu00001 Otu00005 Otu00015 Otu00019 Otu00200 Otu00250 
    1644     1580     1636     1589     2017     2034

Rsq 0.7499362 with otus:
Otu00005 Otu00019 Otu00200 Otu00250 
     758     1121     1421     1700

Rsq 0.7305611 with otus:
Otu00005 Otu00019 
     611      889

Rsq 0.6887173 with otus:
Otu00200 Otu00250 
     550      950

also tried:
Otus	Rsq 
15  5	0.59
250 19	0.757
250 5	0.712

Otu		 Rsq
Otu00001 0.5014594
Otu00005 0.6353671
Otu00015 0.4497236
Otu00019 0.6319819
Otu00200 0.6557830
Otu00250 0.7009930

>>>>

#One per cage
one_mouse_initial_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))

om_raw_otus <- data.frame(otu = colnames(select(one_mouse_initial_df, contains('Otu'))))
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_initial_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_initial_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_raw_otus <- merge(om_raw_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_raw_otus <- select(om_raw_otus, -IncNodePurity)
}
test <- data.frame(om_raw_otus[ , -1], row.names = om_raw_otus$otu)
colnames(test) <- seq(ncol(test))
### rank
test_rank <- apply(test, 2, function(x) rank(x, na.last = F))
test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
test_import_rank <- sort(test_raw_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
### mean
#test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
#test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
#test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
#test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
#### median
#test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
#test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
#test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
#test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)

#####
#test <- c() # test if any model has same train set
#  for(i in 2:21){
#  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
#}
#any(test) # if true, then trained on the same set of mice multiple times
#####
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    filter(!is.na(log_cfu_med))  %>% 
    select(log_cfu_med, mouse_id, cage_id, one_of(names(test_import_rank))))

om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
om_pred <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    om_pred <- rbind(om_pred,
                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                          predicted=predict(rf_om, test)))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_import_otus <- select(om_import_otus, -IncNodePurity)
}
om_rsq <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))

@@

om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
colnames(om_import_otus) <- seq(ncol(om_import_otus))
om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
om_raw_rank <- apply(om_otu_rank, 1, function(x) sum(x, na.rm = T))
med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
                                 importance = om_raw_rank)

@@
